Galois Slicing as Automatic Differentiation

### 2025-07-11 main/models.tex
Obsolete.
[ tex ] tidy up some hanging paragraphs
Some leftmargin parameterisation.
[ tex ] typo

As similar result is given by \citet{cruttwell2022}, where Euclidean
spaces $\RR^n$ and smooth functions are embedded into a category of
lenses (the ``simply typed'' version of the $\Fam$ construction). As
in \citet{vakar22}, the idea is to formally separate functions on
points and their forward/reverse tangent maps for the purposes of
implementation of automatic differentiation. In the case of smooth
maps, this process throws away information on higher derivatives by
turning smooth maps into pairs of plain functions and linear
functions. We conjecture at the end of this section that the analogous
construction in our partially ordered setting does not.

@@

\begin{enumerate}
\item The category $\LatGal$ has bounded lattices as objects and
  Galois connections as morphisms, with the right adjoint going in the
  ``forward'' direction. The category $\Fam(\LatGal)$ is our preferred
  model for total functions with Galois slicing. We explore some
  specific examples in this category in \secref{semantic-gps}.
\item The category $\MeetSLat$ has meet-semilattices with top as
  objects and monotone finite meet preserving functions as
  morphisms. The category $\Fam(\MeetSLat)$ provides a model of
  total functions with ``forward derivatives'' only.
\item The category $\JoinSLat$ has join-semilattices with bottom as
  objects and monotone finite join preserving functions as
  morphisms. The category $\Fam(\JoinSLat^\op)$ provides a model of
  total functions with ``backwards derivatives'' only.
\end{enumerate}

@@

  \begin{salign*}
    (\biinj_X \comp \biproj_X) + (\biinj_Y \comp \biproj_Y) &= \id_{X \biprod Y}
  \end{salign*}
  A zero object is an object that is both initial and terminal.
\end{definition}
As the name suggests, biproducts in a category are both products and
coproducts:
\begin{proposition}
  \item
  \begin{enumerate}
  \item A $\CMon$-category that has biproducts $X \oplus Y$ for all
    $X$ and $Y$ also has products and coproducts with
    $X \times Y = X + Y = X \oplus Y$.
  \item A $\CMon$-category with (co)products also has biproducts, and
    any initial or terminal object is a zero object.
  \end{enumerate}
\end{proposition}

\begin{example}
  The following are $\CMon$-enriched and have finite products, and
  hence biproducts:
  \begin{enumerate}
  \item In $\FinVect$, morphisms are linear maps and so can be added
    and have a zero map. Finite products are given by cartesian
    products of the underlying sets, with the vector operations
    defined pointwise.
  \item In $\LatGal$, right adjoints are summed using meets and left
    adjoints are summed using joins. The zero maps are given by the
    constantly $\top$ and constantly $\bot$ functions
    respectively. Products are given by the cartesian product of the
    underlying set and the one-element lattice for the
    terminal/initial/zero object.
  \item $\MeetSLat$ and $\JoinSLat$ are both $\CMon$-enriched and have
    finite products for similar reasons to $\LatGal$.
  \end{enumerate}
\end{example}

>>>>

A similar result is given by \citet{cruttwell2022}, where Euclidean
spaces $\RR^n$ and smooth functions are embedded into a category of
lenses (the ``simply typed'' version of the $\Fam$ construction). As
in \citet{vakar22}, the idea is to formally separate functions on
points and their forward/reverse tangent maps for the purposes of
implementation of automatic differentiation. In the case of smooth
maps, this process throws away information on higher derivatives by
turning smooth maps into pairs of plain functions and linear
functions. We conjecture at the end of this section that the analogous
construction in our partially ordered setting does not.

@@

\begin{enumerate}[leftmargin=\enummargin]
\item The category $\LatGal$ has bounded lattices as objects and
  Galois connections as morphisms, with the right adjoint going in the
  ``forward'' direction. The category $\Fam(\LatGal)$ is our preferred
  model for total functions with Galois slicing. We explore some
  specific examples in this category in \secref{semantic-gps}.
\item The category $\MeetSLat$ has meet-semilattices with top as
  objects and monotone finite meet preserving functions as
  morphisms. The category $\Fam(\MeetSLat)$ provides a model of
  total functions with ``forward derivatives'' only.
\item The category $\JoinSLat$ has join-semilattices with bottom as
  objects and monotone finite join preserving functions as
  morphisms. The category $\Fam(\JoinSLat^\op)$ provides a model of
  total functions with ``backwards derivatives'' only.
\end{enumerate}

@@

  \begin{salign*}
    (\biinj_X \comp \biproj_X) + (\biinj_Y \comp \biproj_Y) &= \id_{X \biprod Y}
  \end{salign*}
  A zero object is an object that is both initial and terminal.
\end{definition}
As the name suggests, biproducts in a category are both products and
coproducts:
\begin{proposition}
  \item
  \begin{enumerate}[leftmargin=\enummargin]
  \item A $\CMon$-category that has biproducts $X \oplus Y$ for all
    $X$ and $Y$ also has products and coproducts with
    $X \times Y = X + Y = X \oplus Y$.
  \item A $\CMon$-category with (co)products also has biproducts, and
    any initial or terminal object is a zero object.
  \end{enumerate}
\end{proposition}

\begin{example}
  The following are $\CMon$-enriched and have finite products, and
  hence biproducts:
  \begin{enumerate}[leftmargin=\enummargin]
  \item In $\FinVect$, morphisms are linear maps and so can be added
    and have a zero map. Finite products are given by cartesian
    products of the underlying sets, with the vector operations
    defined pointwise.
  \item In $\LatGal$, right adjoints are summed using meets and left
    adjoints are summed using joins. The zero maps are given by the
    constantly $\top$ and constantly $\bot$ functions
    respectively. Products are given by the cartesian product of the
    underlying set and the one-element lattice for the
    terminal/initial/zero object.
  \item $\MeetSLat$ and $\JoinSLat$ are both $\CMon$-enriched and have
    finite products for similar reasons to $\LatGal$.
  \end{enumerate}
\end{example}
